The Phenotypic Consequences of Transposable Elements in Cultivated Strawberry

### 2024-07-31 doc/manuscript/main.tex
Edits made with Jordan 7/31/2024

\section{Background:}
Cultivated garden strawberry (\textit{Fragaria x ananassa} Duchesne ex Rozier, \textit{2n} $=$ 8X $=$ 56) is an allo-octoploid species formed from the hybridization of two New World strawberry species \textit{Fragaria virginiana} and \textit{Fragaria chiloensis}.
It is a major fruit crop; 1,211,090 tonnes were produced in the United States in 2021... TODO.
%(\href{TODO CITE FAOSTAT}{https://data.un.org/Data.aspx?d=FAO&f=itemCode\%3A544}) and the US production value was evaluated as \$3,398,943,000 in 2023 (\href{TODO CITE USDA}{https://www.nass.usda.gov/Statistics_by_Subject/result.php?C6E91360-DC42-3C4A-8D5C-1392239A0EAF&sector=CROPS&group=FRUIT%20%26%20TREE%20NUTS&comm=STRAWBERRIES}.
Its initial hybridization occurred in the 18th century in France, which makes strawberry one of the most recent domesticated crop species.
Unlike most domesticated crops, strawberry has not undergone thousands of years of breedings, only about 300 years of breeding.

However the genomic changes associated with domestication have largely remained a mystery as a high-quality reference genome for cultivated strawberry was not available until quite recently \cite{Edger2019} (TODO CITE RR).
These datasets have enabled a wide-array of studies to be performed on the strawberry genome as several research groups have worked on identifying selective-sweep regions, in addition to surveying the genomic diversity, evolutionary history, and structure of the strawberry genome (TODO CITE).
For some time the genetic variation in \textit{Fragaria x ananassa} was considered to be rather limited (TODO CITE), a notion that has been strongly challenged by recent research (TODO CITE HARDIGAN).
In fact, Hardigan et al showed that genomic diversity is actually quite high, both in the wild founder and domesticated populations.
Other hallmarks of domesticated genomes are absent in strawberry such as ... TODO.
Additionally, the genome shows high degree of karyotypic stability and absence of chromosome restructuring \cite{Hardigan2020, Hardigan2021}.

@@

>>>>

\section{Background:}
Cultivated garden strawberry (\textit{Fragaria x ananassa} Duchesne ex Rozier, \textit{2n} $=$ 8X $=$ 56) is an allo-octoploid species formed from the hybridization of two New World strawberry species \textit{Fragaria virginiana} and \textit{Fragaria chiloensis}.
It is a major fruit crop; 1,211,090 tonnes were produced in the United States in 2021... TODO.
%(\href{TODO CITE FAOSTAT}{https://data.un.org/Data.aspx?d=FAO&f=itemCode\%3A544}) and the US production value was evaluated as \$3,398,943,000 in 2023 (\href{TODO CITE USDA}{https://www.nass.usda.gov/Statistics_by_Subject/result.php?C6E91360-DC42-3C4A-8D5C-1392239A0EAF&sector=CROPS&group=FRUIT%20%26%20TREE%20NUTS&comm=STRAWBERRIES}.
Its initial hybridization occurred in the 18th century in France, which makes strawberry one of the most recent domesticated crop species.
Unlike most domesticated crops, strawberry has not undergone thousands of years of breeding, only about 300 years of breeding. \\

While great strides have been made in a multitude of agriculturally relevant traits such as yield, fruit quality, and disease resistance, the genomic changes associated with strawberry domestication have largely remained a mystery as a high-quality reference genome was not available until recently \cite{Edger2019} (TODO CITE RR, and Hardigan).
(Other datasets like transcriptome/expression atlas?)
This has enabled researchers to quantify the amount of genetic diversity and heterozygosity in the genome, revealing that cultivated strawberry is a highly-diverse genome, as over 70\% of alleles from wild populations were maintained.
Additionally, some hallmarks of selection have been found via the identification of domestication-related selective sweeps as well as QTL associated with fruit quality traits (TODO CITE HARDIGAN 2021a)

While progress has been made investigating the genomic changes associated with domestication, the role of transposable elements (TEs) and the roles they may have played in this process has not been addressed.
TEs are rich sources of genotypic and phenotypic diversity, and have been implicated in a number of plant traits such as 
For example, TEs are implicated in the domestication of \textit{Zea mays} (TODO CITE ROSS tb1), the 
% It is likely that TEs may have played a role in domestication, aside from corn, ....

% Talk about the tradeoff of managing silencing TEs but they also have the potential to create positive change
% Wrap everyting with our appproach, most of the time unique TE-gene instances are the result of a unique phenotype tat is found and then the TE is found to be the causal mutation, here we are starting by looking for TEs and then proceeding from there.

These datasets have enabled a wide-array of studies to be performed on the strawberry genome as several research groups have worked on identifying selective-sweep regions, in addition to surveying the genomic diversity, evolutionary history, and structure of the strawberry genome (TODO CITE).
For some time the genetic variation in \textit{Fragaria x ananassa} was considered to be rather limited (TODO CITE), a notion that has been strongly challenged by recent research (TODO CITE HARDIGAN).
In fact, Hardigan et al showed that genomic diversity is actually quite high, both in the wild founder and domesticated populations.
Other hallmarks of domesticated genomes are absent in strawberry such as ... TODO.
Additionally, the genome shows high degree of karyotypic stability and absence of chromosome restructuring \cite{Hardigan2020, Hardigan2021}.

@@

Whereas previous studies have identified TEs that are responsible for novel traits and potential domestication phenotypes, our approach here is unique because its aim is to identify regions of high TE density within selective sweeps that are near genes associated with domestication phenotypes which may be influenced by TEs.
